Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-13 fscd19/fscd19.tex
[ fscd19 ] squeezing to make room for introduction

While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
normal forms of positive types $P$, their extension $\tcase^B$ and
$\tabort^B$ to negative types
is admissible, for instance:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
  \tabort^{\tyone}     & u & = & \tunit \\
  \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
  \tabort^P           & u & = & \tabort\;u \\
\end{array}
\]
$\tcase$ generalizes analogously, with a bit of care when weakening
the branches.

@@

We will interpret simple types $A$ as corresponding presheaves
$\den A$.  Let us start with the negative types, defining
reflection $\up A : \tNe\, A \todot \den A$
and reification $\down A : \den A \todot \tNf A$ along the way.
\[
\begin{array}{lcl}
  \den   \tyone          & = & \hatone \\
  \upG   \tyone \; u     & = & \ttempty \\
  \downG \tyone \ttempty & = & \tunit \\
\end{array}
\qquad\qquad\qquad
\begin{array}{lcl}
  \den   {A \times B}       & = & \den A \hattimes \den B \\
  \upG   {A \times B} \; u  & = & (\upG A (\prj1 u),\;\upG B (\prj2 u)) \\
  \downG {A \times B} (a,b) & = & \tpair\,(\downG A a)\,(\downG B b) \\
\end{array}
\]
\[
\begin{array}{lcl}
  \den {A \To B}    & = & \den A \hatto \den B \\
  \upG   {A \To B}    \; u \; (\tau : \Gamma \sublist \Delta) \; (a : \den A \Delta)
     & = & \up[\Delta] B (\tapp\;(\tren\,\tau\, u)\;(\down[\Delta] A a)) \\
  \downG {A \To B}    f        & = & \tabs\,(\down[\cext \Gamma A] B
    (f\;\twk^A_\Gamma\;\tfresh^A_\Gamma))
\end{array}
\]
% We will interpret simple types $A$ as corresponding presheaves
% $\den A$.  Let us start with the negative types:
% \[
% \begin{array}{lcl}
%   \den \tyone       & = & \hatone \\
%   \den {A \times B} & = & \den A \hattimes \den B \\
%   \den {A \To B}    & = & \den A \hatto \den B \\
% \end{array}
% \]
% Reflection $\up A : \tNe\, A \todot \den A$
% and reification $\down A : \den A \todot \tNf A$
% are defined mutually by induction on type $A$.
% \begin{gather*}
% \begin{array}{lcl@{\qquad}lcl}
%   \upG   \tyone       u        & = & \ttempty &
%   \downG \tyone       \ttempty & = & \tunit \\
%   \upG   {A \times B} u        & = & (\upG A (\prj1 u),\;\upG B (\prj2 u)) &
%   \downG {A \times B} (a,b)    & = & \tpair\,(\downG A a)\,(\downG B b) \\
% \end{array}
% \\
% \begin{array}{lcl@{\qquad}lcl}
%   \upG   {A \To B}    u \; (\tau : \Gamma \sublist \Delta) \; (a : \den A \Delta)
%      & = & \up[\Delta] B (\tapp\;(\tren\;\tau\; u)\;(\down[\Delta] A a)) \\
%   \downG {A \To B}    f        & = & \tabs\,(\down[\cext \Gamma A] B
%     (f\,(\twk^A)\,(\tfresh^A_\Gamma)))
% \end{array}
% \end{gather*}
%In the equation for
In the reification at function types
$\down{A \To B}$,
the renaming $\twk^A_\Gamma : \Gamma \sublist \cext \Gamma A$
makes room
for a new variable of type $A$, which is reflected into $\den A$ by
\[
\begin{array}{lcl@{\qquad}lcl}
  \tfresh : \den A (\cext \Gamma A) \\
  \tfresh = \up[\cext \Gamma A] A (\tvar\;\tzero)
  . \\
\end{array}
\]
The ability to introduce fresh variables into a context, and to use
semantic objects such as $f : \den{A \To B}\Gamma$ in a such extended
context, is the reason for utilizing presheaves instead of just sets
as semantic types.

@@

Herein the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.

@@

  % \fdent{t : B \from \cext \Gamma A}\gamma & = &
  %   \lambda
  %     .\
  %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\[1ex]
  \dent{\tabs\,t}\gamma
      & = & \fdent t \gamma \\
  % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma \;
  %    (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi) & = &
  % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
  %    \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
  % \dent{\tapp\,(t : A \To B \from \Gamma)\,u} \gamma & = &

@@

@@

  % \dent{\tcase\,(u : A_1 + A_2 \from \Gamma)\;t_1\;t_2}\gamma & = &
    %  \dencase\;\dent u \gamma\;f_1\;f_2
    % \ \bwhere \\
    % \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
    %     & = &
    %     \dent {t_i} {(\tren\;\tau\;\gamma,\ {a_i})}
   %  (\lambda\; (\tau : \Delta \sublist \Phi)\,(b : \den A \Phi).\
   %      \dent v {(\tren\;\tau\;\gamma,\ b)})

@@

@@

@@

Focalization is a technique to remove don't-care non-determinism from
proof search, and as such, polarized lambda calculus is foremost a
calculus of normal forms.
These normal forms are given by four mutually defined inductive
families of presheaves $\tVnf\,P$,
$\tNe\,N$, $\Cov\,\J$,
and $\tNf\,N$.  As they are very similar to the CBPV normal forms
given in the last section we only report the differences.

Values \fbox{$v : \PVal P \Gamma$} are unchanged, they can refer to
atomic positive hypotheses ($\pvar$) and normal $\tthunk$s.

Neutrals \fbox{$\Ne N \Gamma$} start with a negative variable instead
of with $\tforce$, as forcing thunks is already performed in $\nhyp$
when adding analysing hypotheses of $\tThunk$ type.
\[
  \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
\]

The Cover monad \fbox{$\Cover \J \Gamma$}
lacks constructors $\tsplit$, $\tcase$ and $\tabort$
since the pattern matching is taken care of by $\Add P$.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
      {\Cover \J \Gamma}
\end{gather*}

The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
with the exception that pattern matching happens eagerly in $\tabs$.
\[
  \nru{\tabs}
      {\AddHypP P {\tNf\, N} \Gamma}
      {\NfP {P \To N} \Gamma}
\]

@@

The method $\reflect P$ replaces previous uses of $\tfresh^P$ in
reflection and reification at negative types.
% and identity environment creation.
\[
\begin{array}{l@{~}lcl}
  \downG {P \To N} & f & = & \tabs
    \left(
      \reflect[\Gamma] P
        \left(
          \lambda \, (\tau : \Gamma \sublist \Delta) \; a .\
          \down[\Delta] P (f \; \tau \; a)
        \right)
    \right)
  \\
  \upG {\Comp P} & u & = & \tbind \; u
    \left(
      \reflect[\Gamma] P
        \left(
          \lambda \; \tau \; a .\
          \treturn\; a
        \right)
    \right)
  % \\
  % \idenv{\cext \Gamma P} & = &
\end{array}
\]
% To create an identity environment using $\treflect$, let us first
% extend $\Add\_$ to contexts, following the intuition that
Due to the strong limitation on context entries,
the identity environment $\idenv \Gamma$ can be built
straightforwardly using negative reflection.
\[
\begin{array}{lcl}
  \idenv{\cempty} & = & \ttempty \\
  \idenv{\cext \Gamma {\patom}} & = &
    \left(
      \tren\;\twk^{\patom}\;\idenv\Gamma
      , \
      \pvar\,\tzero
    \right) \\
  \idenv{\cext \Gamma N} & = &
    \left(
      \tren\;\twk^N\;\idenv\Gamma
      , \
      \upG N (\nvar\,\tzero)
    \right) \\
\end{array}
\]

>>>>

While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
normal forms of positive types $P$, their extension $\tcase^B$ and
$\tabort^B$ to negative types
is admissible, for instance:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
  \tabort^{\tyone}     & u & = & \tunit \\
  \tabort^P           & u & = & \tabort\;u \\
\end{array}
\qquad
\begin{array}{l@{~}lcl}
  \\
  \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
\end{array}
% \begin{array}{l@{~}lcl}
%   \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
%   \tabort^{\tyone}     & u & = & \tunit \\
%   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
%   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
%   \tabort^P           & u & = & \tabort\;u \\
% \end{array}
\]
$\tcase$ generalizes analogously, with a bit of care when weakening
the branches.

@@

We will interpret simple types $A$ as corresponding presheaves
$\den A$.  Let us start with the negative types, defining
reflection $\up A : \tNe\, A \todot \den A$
and reification $\down A : \den A \todot \tNf A$ along the way.
\[
\begin{array}{lcl}
  \den   \tyone          & = & \hatone \\
  \upG   \tyone \; u     & = & \ttempty \\
  \downG \tyone \ttempty & = & \tunit \\
\end{array}
\qquad\qquad\qquad
\begin{array}{lcl}
  \den   {A \times B}       & = & \den A \hattimes \den B \\
  \upG   {A \times B} \; u  & = & (\upG A (\prj1 u),\;\upG B (\prj2 u)) \\
  \downG {A \times B} (a,b) & = & \tpair\,(\downG A a)\,(\downG B b) \\
\end{array}
\]
\[
\begin{array}{lcl}
  \den {A \To B}    & = & \den A \hatto \den B \\
  \upG   {A \To B}    \; u \; (\tau : \Gamma \sublist \Delta) \; (a : \den A \Delta)
     & = & \up[\Delta] B (\tapp\;(\tren\,\tau\, u)\;(\down[\Delta] A a)) \\
  \downG {A \To B}    f        & = & \tabs\,(\down[\cext \Gamma A] B
    (f\;\twk^A_\Gamma\;\tfresh^A_\Gamma))
\end{array}
\]
% We will interpret simple types $A$ as corresponding presheaves
% $\den A$.  Let us start with the negative types:
% \[
% \begin{array}{lcl}
%   \den \tyone       & = & \hatone \\
%   \den {A \times B} & = & \den A \hattimes \den B \\
%   \den {A \To B}    & = & \den A \hatto \den B \\
% \end{array}
% \]
% Reflection $\up A : \tNe\, A \todot \den A$
% and reification $\down A : \den A \todot \tNf A$
% are defined mutually by induction on type $A$.
% \begin{gather*}
% \begin{array}{lcl@{\qquad}lcl}
%   \upG   \tyone       u        & = & \ttempty &
%   \downG \tyone       \ttempty & = & \tunit \\
%   \upG   {A \times B} u        & = & (\upG A (\prj1 u),\;\upG B (\prj2 u)) &
%   \downG {A \times B} (a,b)    & = & \tpair\,(\downG A a)\,(\downG B b) \\
% \end{array}
% \\
% \begin{array}{lcl@{\qquad}lcl}
%   \upG   {A \To B}    u \; (\tau : \Gamma \sublist \Delta) \; (a : \den A \Delta)
%      & = & \up[\Delta] B (\tapp\;(\tren\;\tau\; u)\;(\down[\Delta] A a)) \\
%   \downG {A \To B}    f        & = & \tabs\,(\down[\cext \Gamma A] B
%     (f\,(\twk^A)\,(\tfresh^A_\Gamma)))
% \end{array}
% \end{gather*}
%In the equation for
In the reification at function types
$\down{A \To B}$,
the renaming $\twk^A_\Gamma : \Gamma \sublist \cext \Gamma A$
makes room
for a new variable of type $A$, which is reflected into $\den A$ by
\fbox{$\tfresh^A_\Gamma = \up[\cext \Gamma A] A {\vv0} : \den A (\cext \Gamma A)$}.
% \[
% \begin{array}{lcl@{\qquad}lcl}
%   \tfresh : \den A (\cext \Gamma A) \\
%   \tfresh = \up[\cext \Gamma A] A {\vv0}
%   . \\
% \end{array}
% \]
The ability to introduce fresh variables into a context, and to use
semantic objects such as $f : \den{A \To B}\Gamma$ in a such extended
context, is the reason for utilizing presheaves instead of just sets
as semantic types.

@@

Herein, the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.

@@

@@

  \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\

@@

@@

%   \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
%   \dent{\tunit}\gamma & = & \ttempty \\
%   \dent{\tpair\,t_1\,t_2}\gamma & = &
%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
%   \dent{\tapp\,t\,u} \gamma & = &
%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
%   \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
%   \dent{\tapp\,t\,u} \gamma & = &
%   \dent{\tunit}\gamma & = & \ttempty \\
%   \dent{\tpair\,t_1\,t_2}\gamma & = &
%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\

@@

% \[
%   % \fdent{t : B \from \cext \Gamma A}\gamma & = &
%   %   \lambda
%   %     \,(\tau : \Delta \sublist \Phi)
%   %     \,(a : \den A \Phi)
%   %     .\
%   %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\[1ex]
%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
%   \dent{\tabs\,t}\gamma
%       & = & \fdent t \gamma \\
%   % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma \;
%   %    (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi) & = &
%   %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
%   % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
%   %    \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
%   %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
%   \dent{\tapp\,t\,u} \gamma & = &
%   % \dent{\tapp\,(t : A \To B \from \Gamma)\,u} \gamma & = &
%   %    \dent t \gamma\;\tid\;\dent u \gamma \\
%   \dent{\tunit}\gamma & = & \ttempty \\
%   \dent{\tpair\,t_1\,t_2}\gamma & = &
%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
%   % \dent{\tcase\,(u : A_1 + A_2 \from \Gamma)\;t_1\;t_2}\gamma & = &
%     %  \dencase\;\dent u \gamma\;f_1\;f_2
%     % \ \bwhere \\
%     % \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
%     %     & = &
%     %     \dent {t_i} {(\tren\;\tau\;\gamma,\ {a_i})}
%    % \
%    %  (\lambda\; (\tau : \Delta \sublist \Phi)\,(b : \den A \Phi).\
%    %      \dent v {(\tren\;\tau\;\gamma,\ b)})
% \]

@@

Focalization is a technique to remove don't-care non-determinism from
proof search, and as such, polarized lambda calculus is foremost a
calculus of normal forms.
These normal forms are given by four mutually defined inductive
families of presheaves $\tVnf\,P$,
$\tNe\,N$, $\Cov\,\J$,
and $\tNf\,N$.  As they are very similar to the CBPV normal forms
given in the last section, we only report the differences.
%
Values \fbox{$v : \PVal P \Gamma$} are unchanged, they can refer to
atomic positive hypotheses ($\pvar$) and normal $\tthunk$s.
%
Neutrals \fbox{$\Ne N \Gamma$} start with a negative variable instead
of with $\tforce$, as forcing thunks is already performed in $\nhyp$
when adding analysing hypotheses of $\tThunk$ type.
% \[
%   \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
% \]
%
The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
with the exception that pattern matching happens eagerly in $\tabs$.
\[
  \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
\qquad
\qquad
  \nru{\tabs}
      {\AddHypP P {\tNf\, N} \Gamma}
      {\NfP {P \To N} \Gamma}
\]
%
The Cover monad \fbox{$\Cover \J \Gamma$}
lacks constructors $\tsplit$, $\tcase$ and $\tabort$
since the pattern matching is taken care of by $\Add P$.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
      {\Cover \J \Gamma}
\end{gather*}

@@

The method $\reflect P$ replaces previous uses of $\tfresh^P$ in
reflection and reification at negative types.
% and identity environment creation.
\[
\begin{array}{l@{~}lcl}
  \downG {P \To N} & f & = & \tabs
    \left(
      \reflect[\Gamma] P
        \left(
          \lambda \, (\tau : \Gamma \sublist \Delta) \; a .\
          \down[\Delta] P (f \; \tau \; a)
        \right)
    \right)
  \\
  \upG {\Comp P} & u & = & \tbind \; u
    \left(
      \reflect[\Gamma] P
        \left(
          \lambda \; \tau \; a .\
          \treturn\; a
        \right)
    \right)
  % \\
  % \idenv{\cext \Gamma P} & = &
\end{array}
\]
% To create an identity environment using $\treflect$, let us first
% extend $\Add\_$ to contexts, following the intuition that
Due to the strong limitation on context entries,
the identity environment $\idenv \Gamma$ can be built
straightforwardly using negative reflection.
\[
\begin{array}{lcl}
  \idenv{\cempty} & = & \ttempty \\
  \idenv{\cext \Gamma {\patom}} & = &
    (
      \tren\;\twk^{\patom}\;\idenv\Gamma
      , \
      \pvar\,\tzero
    ) \\
  \idenv{\cext \Gamma N} & = &
    (
      \tren\;\twk^N\;\idenv\Gamma
      , \
      \upG N (\nvar\,\tzero)
    ) \\
\end{array}
% \begin{array}{lcl}
%   \idenv{\cempty} & = & \ttempty \\
%   \idenv{\cext \Gamma {\patom}} & = &
%     \left(
%       \tren\;\twk^{\patom}\;\idenv\Gamma
%       , \
%       \pvar\,\tzero
%     \right) \\
%   \idenv{\cext \Gamma N} & = &
%     \left(
%       \tren\;\twk^N\;\idenv\Gamma
%       , \
%       \upG N (\nvar\,\tzero)
%     \right) \\
% \end{array}
\]
